Estimating duration in the presence of misclassification

### 2024-01-26 cis-imperfect-testing.tex

@@

Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}.
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having dense data collection here.
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. CIS-based has more episodes ending in the first 14 days. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_i = e_i - b_i + 1 = 1$.
Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
\todo{possibly a bit unclear phrasing here? struggling with it}
This produced the same pattern, that is estimating too many short episodes (not shown).

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are two strands of evidence supporting this hypothesis.
First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Second, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see \cref{imperf-test:sec:simulate} for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

A better model would be to allow the rate of false negatives to vary over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

@@

Now consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
I proceed by considering whether this is the case.

First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
These occur with probability 1, by the assumption of no false positives.
The tests in $\sched_i'$ are either true positives or false negatives.
\begin{align}
&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\end{align}

Second, if $e_i > r_i^{(e)}$.
In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
As I only consider the tests in $\sched_i'$, the end of the episode is right-censored.
Taking the same approach as before:
\begin{align}
&p(\vec{y_i'}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
\end{align}

Combining the above, the replacement for $p_{ia}$ is:
\begin{align}
p_{ia}'
&= p(\vec{y_i'} \mid p_\text{sens}, \theta) \\
&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(\vec{y_i'}, b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.

@@

This false negative would occur on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is $d_i > \tau_{\sched_i}(b_i)$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

@@

The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_i) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_i) + 1
    \mid b_i,
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as \cref{E-perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$.
For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

>>>>

{\color{red}{NEW BIT}}
In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
However, initial application of this framework produced implausibly short estimate. In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).  In \cref{imperf-test:sec:modelling} I develop the model in \cref{E-perf-test} to include the possibility of observing false negatives.
There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
\todo[inline]{Review literature properly, here or elsewhere}. Building on this literature, model adaptation to include false negatives involves.....GIVE DETAILS OF WHAT YOU DID
After showing through simulation (\cref{imperf-test:sec:sim-study-results}) that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}),  and I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

Applying the method of \cref{E-perf-test} to CIS data estimates fewer individuals of a duration less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having dense data collection here.
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. CIS-based has more episodes ending in the first 14 days. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_i = e_i - b_i + 1 = 1$.
Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
% This produced the same pattern, that is estimating too many short episodes (not shown).

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

A better model would be to allow the rate of false negatives to vary over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

@@

Now consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
I proceed by considering whether this is the case, and considering the case with $b_i$ known; $b_i$ will later be integrated out.

First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
These occur with probability 1, by the assumption of no false positives.
The tests in $\sched_i'$ are either true positives or false negatives.
\begin{align}
&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, if $e_i > r_i^{(e)}$.
In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_i^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_i^{(e)}$.
\todo{check previous notes - I think I investigated this assumption a bit more and maybe did some sensitivity analysis}
Taking the same approach as before:
\begin{align}
&p(\vec{y_i'}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_i'}$.
First, augment the data with $b_i$, and split into the cases just discussed:
\begin{align}
p_{ia}'
&= p(\vec{y_i'} \mid p_\text{sens}, \theta) \\
&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, e_i \leq r_i^{(e)} \mid b_i, p_\text{sens}, \theta) + p(\vec{y_i'}, e_i > r_i^{(e)} \mid b_i, p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta).
\intertext{This simplifies to:}
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.

@@

This false negative would occur at the first test after the infection episode begins, on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_i)$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

@@

The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_i) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_i) + 1
    \mid b_i,
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_i$ for $\min \sched_i < b_i \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
$p_{iu}'$ is the probability of episode $i$ being missed, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
These mechanisms are mutually exclusive.
Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

### 2024-01-26 incidence-prevalence.tex

\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

>>>>

\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
\todo{add discussion of proportions vs rates}
As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

### 2024-01-26 transmission.tex

\todo[inline]{
    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
    At the same place, I could review the literature on mechanistic models using REACT and CIS.
}

@@

Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

>>>>

\todo[inline]{
    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}. Also Toffe.
    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
    At the same place, I could review the literature on mechanistic models using REACT and CIS.
}

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
